A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-11-11 inst/manuscript/manuscript.tex
minor updates

The design-based approach can be used to select samples that are
``well-spread'' in space, or spatially balanced. Spatially balanced
samples are useful because parameter estimates from these samples tend
to vary less than parameter estimates from samples that are not
spatially balanced (Barabesi and Franceschi, 2011; Benedetti et al.,
2017; Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004). To
quantify the spatial balance of a sample, Stevens and Olsen (2004)
proposed loss metrics based on Voroni polygons. After the GRTS algorithm
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use the Generalized Random Tessellation Stratified
(GRTS) algorithm to select spatially balanced samples sampling because
the algorithm has several attractive properties. It accommodates finite
and infinite sampling frames. It accommodates equal, unequal, and
proportional (to size) inclusion probabilities. It accommodates legacy
(historical) sampling (Foster et al., 2017). It accommodates a minimum
distance between units in a sample. It accommodates replacement units in
a sample, which are units that can be sampled in place of an original
unit that can no longer be sampled. Lastly, it can be implemented using
the spsurvey R package Dumelle et al. (2021).

The GRTS algorithm samples from finite and infinite populations by
utilizing a mapping between two-dimensional and one-dimensional space.
The units in the two-dimensional sampling frame are divided into cells
using a hierarchical address. This hierarchical address is then used to
map the units from two-dimensional space to a one-dimensional line where
each unit's line length equals its inclusion probability. A systematic
sample is conducted on the line and linked back to a unit in
two-dimensional space, which results in the desired sample. Stevens and
Olsen (2004) provides further details.

@@

Let \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for teh
sampled and unsampled population units, respectively. We assume
\(E(\bm{\delta} = \mathbf{0}\) and that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th elment of hte matrix
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is dependent random error variance
measuring coarse-scale (correlated) variability, \(\sigma^2_{n}\) is the
independent random error variance measuring fine-scale (independent)
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the correlation, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). Often
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are called the partial sill and
nugget, respectively. Any spatial covariance function could be used in
the place of the exponential, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

@@

We used a numerical simulation study to investigate performance of four
design-analysis combinations, summarized in Table
\ref{tab:designanalysis}.

\begin{table}[ht]
\centering
\begin{tabular}{r|ll}
  \hline
 & Design & Model \\ 
  \hline
IRS & IRS-Design & IRS-Model \\ 
  GRTS & GRTS-Design & GRTS-Model \\ 
   \hline
\end{tabular}
\caption{\label{tab:designanalysis} Types of Sampling Design and Analysis combinations considered in the simulation study. The rows give the two types of sampling designs while the columns give the two types of analyses.} 
\end{table}

We used a crossed design with the simulation parameters given in Table
\ref{tab:parmtab} for a total of 36 scenarios. All scenarios used
exponential correlation with a \(\sqrt{4} / 3\) for \(N = 900\) response
values simulated on the unit square in either random locations (Layout =
Random) or gridded locations (Layout = Gridded). The mean for the
spatial process generating the response was set to zero.

@@

\begin{table}[ht]
\centering
\begin{tabular}{r|lll}
   \hline
Sample Size (n) & 50 & 100 & 200 \\ 
  Layout & Random & Gridded & - \\ 
  Proportion of Dependent Error & 0 & 0.5 & 0.9 \\ 
  Response Type & Normal & Lognormal & - \\ 
   \hline
\end{tabular}
\caption{\label{tab:parmtab} Simulation parameters. Total variability for all scenarios was 2.} 
\end{table}

@@

The GRTS algorithm and the local neighborhood variance estimator are
available in the \textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle
et al., 2021). FPBK can be readily performed in \texttt{R} with the
\texttt{sptotal} package (Higham et al., 2021). We use \texttt{sptotal}
for both the simulation analysis and the application, estimating
parameters with Restricted Maximum Likelihood (REML).

Figure \ref{fig:figeff} shows the relative efficiency of the four
approaches from Table \ref{tab:designanalysis} using the random location
layout, where ``IRS-Design'' is the baseline: \mbox{} \begin{equation*}
\text{EFF} = \frac{\text{rMS(P)E of approach}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

where rMS(P)E is the root-mean-squared error (design-based) or the
root-mean-squared-prediction error (model-based). When there is no
spatial correlation (top row), the four approaches have approximately
equal rMSPE, even when the assumptions of the model-based approaches are
violated. So, using GRTS or using a spatial model does not result in
much, if any, loss in efficiency even if the response variable is not
spatially correlated. When there is high spatial correlation (bottom
row), the GRTS-Model approach tends to perform best, but difference in
relative efficiency between GRTS-Model and GRTS-Design is small. In the
lognormal, high partial sill settings (bottom-right facet), GRTS-Design
outperforms IRS-Model by a large margin, suggesting that the design
decision (whether to use IRS or GRTS) is more important than the
analysis decision (whether to analyze using model assumptions or not).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/figeff-1} \caption{Relative Efficiency of the four design-analysis approaches. The plot is faceted by the type of response on the columns and the partial-sill to total-variance ratio on the rows.}\label{fig:figeff}
\end{figure}

We also studied 95\% interval coverage among the approaches. The
design-based 95\% confidence intervals and model-based 95\% prediction
intervals are constructed using the normal distribution. Justification
for the design-based intervals lies in the asymptotic normality of
totals via the Central Limit Theorem, and justification for the
model-based intervals lies in the normality assumption of the errors.
Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four approaches in the random location layout. All four approaches
have somewhat similar interval coverage in all settings, with
GRTS-Design having slightly lower coverage when the response is normal.

In the normal response settings, all approaches have coverage around
95\%. This is expected, as the intervals are also based on the normal
distribution In the lognormal response settings, however, all approaches
have coverage below 95\%. This is also expected, as the intervals are
still based on the normal distribution. In the lognormal response
settings, interval coverage increases both as the sample size increases
and as the strength of spatial dependence increases. This suggests that
the larger the sample size and the stronger the spatial dependence, the
more resistant these intervals are to departures from normality of the
data. These conclusions were similar to those observed when the data
were gridded.

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/figconf-1} \caption{Coverage of the four design-analysis approaches. All confidence intervals are normal-based and have a nominal confidence level of 0.95, marked with a horizontal line. The plot is faceted by the type of response on the columns and the partial-sill to total-variance ratio on the rows.}\label{fig:figconf}
\end{figure}

>>>>

The design-based approach can be used to select samples that are
``well-spread'' in space, or spatially balanced. Spatially balanced
samples are useful because parameter estimates from these samples tend
to vary less than parameter estimates from samples that are not
spatially balanced (Barabesi and Franceschi, 2011; Benedetti et al.,
2017; Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004). To
quantify the spatial balance of a sample, Stevens and Olsen (2004)
proposed loss metrics based on Voroni polygons. After the GRTS algorithm
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use the Generalized Random Tessellation Stratified
(GRTS) algorithm to select spatially balanced samples sampling because
the algorithm has several attractive properties. It accommodates finite
and infinite sampling frames. It accommodates equal, unequal, and
proportional (to size) inclusion probabilities. It accommodates legacy
(historical) sampling (Foster et al., 2017). It accommodates a minimum
distance between units in a sample. Lastly, it accommodates replacement
units in a sample, which are units that can be sampled in place of an
original unit that can no longer be sampled. The GRTS algorithm samples
from finite and infinite populations by utilizing a mapping between
two-dimensional and one-dimensional space. The units in the
two-dimensional sampling frame are divided into cells using a
hierarchical address. This hierarchical address is then used to map the
units from two-dimensional space to a one-dimensional line where each
unit's line length equals its inclusion probability. A systematic sample
is conducted on the line and linked back to a unit in two-dimensional
space, which results in the desired sample. Stevens and Olsen (2004)
provides further details.

@@

Let \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for teh
sampled and unsampled population units, respectively. We assume
\(E(\bm{\delta} = \mathbf{0}\) and that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th elment of hte matrix
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is dependent random error variance
measuring coarse-scale (correlated) variability, \(\sigma^2_{2}\) is the
independent random error variance measuring fine-scale (independent)
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the correlation, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). Often
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are called the partial sill and
nugget, respectively. Any spatial covariance function could be used in
the place of the exponential, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations: IRS-Design, IRS with a design-based
analysis; IRS-Model, IRS with a model-based analysis; GRTS-Design, GRTS
sampling with a design-based analysis; and GRTS-Model, GRTS sampling
with a model-based analysis. These combinations are also provided in
Table \ref{tab:designanalysis}.

\begin{table}[ht]
\centering
\begin{tabular}{r|ll}
  \hline
 & Design & Model \\ 
  \hline
IRS & IRS-Design & IRS-Model \\ 
  GRTS & GRTS-Design & GRTS-Model \\ 
   \hline
\end{tabular}
\caption{\label{tab:designanalysis} Sampling-analysis combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.} 
\end{table}

Performance of the four sampling-analysis combinations were evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts, two response
types, and three proportions of dependent random error. The three sample
sizes (\(n\)) were \(n = 50, n = 100,\) and \(n = 200\). Samples were
always selected from a population size (\(N\)) of \(N = 900\). The two
location layouts were random and gridded. Locations in the random layout
were selected randomly from the unit square (\([0, 1] \times [0, 1]\)).
Locations in the gridded layout were selected randomly on a fixed grid
from the unit square. The two response types were normal and lognormal.
For the normal response type, the response was simulated using mean-zero
random errors with the exponential covariance
(Equation\(~\)\ref{equation:expcov}) for varying proportions of
dependent random error. The proportion of dependent random error is
represented by \(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where
\(\sigma^2_1\) and \(\sigma^2_2\) are from
Equation\(~\)\ref{equation:expcov}. The total variance
(\(\sigma^2_1 + \sigma^2_2\)) was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decays to nearly zero at the largest possible distance
between two units in the domain. For the lognormal response type, the
response was first simulated using the same approach as for the normal
response type, except that the total variance was 0.6931 instead of 2.
The response was then exponentiated, yielding a total variance of 2 on
the exponentiated scale. The lognormal responses were used to evaluate
performance of the sampling-analysis approaches for data that were
skewed.

a crossed design with the simulation parameters given in Table
\ref{tab:parmtab} for a total of 36 scenarios. All scenarios used
exponential correlation with a \(\sqrt{4} / 3\) for \(N = 900\) response
values simulated on the unit square in either random locations (Layout =
Random) or gridded locations (Layout = Gridded). The mean for the
spatial process generating the response was set to zero.

@@

\begin{table}[ht]
\centering
\begin{tabular}{r|lll}
   \hline
Sample Size (n) & 50 & 100 & 200 \\ 
  Location Layout & Random & Gridded & - \\ 
  Proportion of Dependent Error & 0 & 0.5 & 0.9 \\ 
  Response Type & Normal & Lognormal & - \\ 
   \hline
\end{tabular}
\caption{\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the mean was zero and the variance was two.} 
\end{table}

@@

The GRTS algorithm and the local neighborhood variance estimator are
available in the \textbf{\textsf{R}} package spsurvey (Dumelle et al.,
2021). FPBK can be readily performed using the \texttt{sptotal}
\textbf{\textsf{R}} package (Higham et al., 2021). We use sptotal for
both the simulation analysis and the application, estimating covariance
parameters with Restricted Maximum Likelihood (REML).

We define rMS(P)E as the root-mean-squared error (design-based) or the
root-mean-squared-prediction error (model-based).Figure \ref{fig:figeff}
shows the relative rMS(P)E of the four approaches from Table
\ref{tab:designanalysis} using the random location layout with
``IRS-Design'' is the baseline. More formally, the relative rMS(P)E is
defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*}

When there is no spatial correlation (Figure \ref{fig:figeff}, top row),
the four approaches have approximately equal rMSPE, even when the
assumptions of the model-based approaches are violated. So, using GRTS
or using a spatial model does not result in much, if any, loss in
efficiency even if the response variable is not spatially correlated.
When there is high spatial correlation (Figure \ref{fig:figeff}, bottom
row), the GRTS-Model approach tends to perform best, but difference in
relative efficiency between GRTS-Model and GRTS-Design is small. In the
lognormal, high partial sill settings (Figure \ref{fig:figeff}, bottom
row and right column), IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/figeff-1} \caption{Relative rMS(P)E for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}\label{fig:figeff}
\end{figure}

We also studied 95\% interval coverage among the approaches. The
design-based 95\% confidence intervals and model-based 95\% prediction
intervals were constructed using the normal distribution. Justification
for the design-based and model-based intervals comes from the asymptotic
normality of totals via the Central Limit Theorem (under some
assumptions).

Figure \ref{fig:figconf} shows the 95\% interval coverage for each of
the four approaches in the random location layout. All four approaches
have somewhat similar interval coverage in all settings, with
GRTS-Design having slightly lower coverage when the response is normal.
Coverage in the normal response settings tended to be near 95\%,
slightly higher than coverage in the lognormal settings, which still
generally exceeded 90\%. Coverage for all approaches increased with the
sample size for both the normal and lognormal scenarios. At a sample
size of 200, all four sampling-design combinations had approximately
95\% interval coverage in both response scenarios. These conclusions
were similar to those observed when the data were gridded.

\textbackslash begin\{figure\}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/figconf-1}
\textbackslash caption\{Interval coverage for the four sampling-analysis
combinations. The rows indicate the proportion of dependent error and
the columns indicate the response type. The solid, black line in each
plot represents 95\% coverage.\}\label{fig:figconf}
\textbackslash end\{figure\}
